---
title: "Project 2"
author: "Adeel, Sadik, Gavin"
date: "7/23/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}

#libraries needed for Predicting Earning Potenital
library(naniar)
library(tidyverse)
library(curl)
library(class)
library(e1071)
library(caret)
library(caret)
library(dplyr)
library(ggthemes)
library(ggplot2)
library(GGally)
library(corrplot)
library(readr)
library(magrittr)
library(ISLR)
library(MASS)
library(FNN)
library(tibble)
library(kdensity)
library(ROCR)
library(VIM)
library(missForest)

```


```{r}

adult <- read.table("C:/Users/Adeel/OneDrive/Desktop/SMU/6372/Project 2 Details 2021/adult.data",sep = ",", fill = F, strip.white=T)

colnames(adult) <- c('age', 'workclass', 'fnlwgt', 'education', 
                     'education_num', 'marital_status', 'occupation', 'relationship', 'race', 'sex', 
                     'capital_gain', 'capital_loss', 'hours_per_week', 'native_country', 'income')
#dataframe statistics, strutcture and analysis
str(adult)
range(adult$age)
dim(adult)

# dealing with "?" observed in the data in workclass, for thsi we will chnage it to NA values and remove it
table(adult$workclass)
adult[adult == "?"] <- NA
table(adult$workclass)

```

combing working classification factors into a smaller group
```{r}
#Checking work calss variables: creating factors: combining and grouping the data: "Never Worked" nad "Without pay" into other/uknown: Federal, state and local to goverenment groups, and self employed groups.
# Combining all workcalss factors
adult$workclass <- as.character(adult$workclass)
 adult$workclass[adult$workclass == "Without-pay" | 
                  adult$workclass == "Never-worked"] <- "Jobless"
 adult$workclass[adult$workclass == "State-gov" |
                  adult$workclass == "Local-gov"]  <- "govt" 
 adult$workclass[adult$workclass == "Self-emp-inc" |
                  adult$workclass == "Self-emp-not-inc"]  <- "Self-employed" 
table(adult$workclass)


# Combining mairtal status
#Chceking data before combining
table(adult$marital_status)
#Combining
adult$marital_status <- as.character(adult$marital_status)
  adult$marital_status[adult$marital_status == "Married-AF-spouse" |
                       adult$marital_status == "Married-civ-spouse" |
                       adult$marital_status == "Married-spouse-absent"] <- "Married"
  adult$marital_status[adult$marital_status == "Divorced" |
                       adult$marital_status == "Separated" |
                       adult$marital_status == "Widowed"] <- "Not-Married"
table(adult$marital_status)

# Combining countries
adult$native_country <- as.character(adult$native_country)
north.america <- c("Canada", "Cuba", "Dominican-Republic", "El-Salvador", "Guatemala",
                   "Haiti", "Honduras", "Jamaica", "Mexico", "Nicaragua",
                   "Outlying-US(Guam-USVI-etc)", "Puerto-Rico", "Trinadad&Tobago",
                   "United-States")
asia <- c("Cambodia", "China", "Hong", "India", "Iran", "Japan", "Laos",
          "Philippines", "Taiwan", "Thailand", "Vietnam")
south.america <- c("Columbia", "Ecuador", "Peru")
europe <- c("England", "France", "Germany", "Greece", "Holand-Netherlands",
            "Hungary", "Ireland", "Italy", "Poland", "Portugal", "Scotland",
            "Yugoslavia")
other <- c("South", "?")
 adult$native_country [adult$native_country %in% north.america] <- "North-America"
 adult$native_country [adult$native_country %in% asia]  <- "Asia"
 adult$native_country [adult$native_country %in% south.america] <- "South-America" 
 adult$native_country [adult$native_country %in% europe] <-  "Europe"  
 adult$native_country [adult$native_country %in% other] <- "Other"
table(adult$native_country)




```

exploring the dataset
```{r}
#change income variable to numerical
#adult$income <- as.numeric(as.factor(adult$income))


#correlation plot
#num.var <- c(1,3,5,11:13,15)
#corrplot(cor(adult[,num.var]))




```

```{r}
# Removing non-self explanatory variables
adult$education <- NULL
adult$fnlwgt <- NULL
adult$relationship <- NULL

```


preprocess continuous variables
```{r}

ggplot(adult, aes(age, fill = income)) + 
  geom_boxplot()+coord_flip() +
  ggtitle("Boxplot for Age by income") +
  theme(plot.title = element_text(hjust = 0.5))

ggplot(adult, aes(age, fill = marital_status)) + 
  geom_boxplot()+coord_flip() +
  ggtitle("Boxplot for Age by marital status") +
  theme(plot.title = element_text(hjust = 0.5))

ggplot(adult, aes(age, fill = workclass)) + 
  geom_boxplot()+coord_flip() +
  ggtitle("Boxplot for Age by Workclass") +
  theme(plot.title = element_text(hjust = 0.5))


ggplot(adult, aes(age, fill = marital_status)) + 
  geom_boxplot()+coord_flip() +
  ggtitle("Boxplot for Age by marital status") +
  theme(plot.title = element_text(hjust = 0.5))


ggplot(adult, aes(education_num, fill = race)) + 
  geom_boxplot()+coord_flip() +
  ggtitle("Boxplot for Education_num by race") +
  theme(plot.title = element_text(hjust = 0.5))

ggplot(adult, aes(capital_loss, fill = race)) + 
  geom_boxplot()+coord_flip() +
  ggtitle("Boxplot for Capital Loss by race") +
  theme(plot.title = element_text(hjust = 0.5))
  
ggplot(adult, aes(capital_gain, fill = race)) + 
  geom_boxplot()+coord_flip() +
  ggtitle("Boxplot for Capital Gain by race") +
  theme(plot.title = element_text(hjust = 0.5))
  
ggplot(adult, aes(hours_per_week, fill = marital_status)) + 
  geom_boxplot()+coord_flip() +
  ggtitle("Boxplot for hours_per_week by marital status") +
  theme(plot.title = element_text(hjust = 0.5))

ggplot(adult, aes(x = education_num, fill = income)) + geom_bar(position="fill") + theme(axis.text.x = element_text(angle = 90)) + ggtitle("Years of Education")

adultCG <- adult %>% group_by(race) %>% summarize(sumcg = sum(capital_gain)) %>% mutate(Freq = round(sumcg/sum(sumcg),3))
adultCG

adultCL <- adult %>% group_by(race) %>% summarize(sumcg = sum(capital_loss)) %>% mutate(Freq = round(sumcg/sum(sumcg),3))
adultCL

ggplot(adult, aes(capital_gain, fill = income)) + 
  geom_histogram() + 
  ggtitle("Boxplot for Capital Gain by income") +
  theme(plot.title = element_text(hjust = 0.5))

ggplot(adult, aes(capital_loss, fill = income)) + 
  geom_histogram() + 
  ggtitle("Boxplot for Capital Loss by income") +
  theme(plot.title = element_text(hjust = 0.5))

ggplot(adult, aes(workclass, fill = income)) + 
  geom_bar() + 
  ggtitle("Boxplot for Work Class by income") +
  theme(plot.title = element_text(hjust = 0.5))


```


```{r}
#missing data analysis
n_miss(adult)
n_complete(adult)

miss_var_summary(adult)

vis_miss(adult)
vis_miss(adult, sort_miss = TRUE)

library(Amelia)
missmap(adult, main = "Missing values")
table (complete.cases(adult))

sum(is.na(adult))


```

viewing type of factors that are available and clean up if required
```{r}
unique(adult$workclass)
unique(adult$marital_status)
unique(adult$occupation)
unique(adult$race)
unique(adult$native_country)


```

taking care of NA's
```{r}
# Checking omition of NA values
adult <- na.omit(adult)
#Missing values you can see that adult data has zero missing values, will use thsi for anlaysis
sum(is.na(adult))


```


```{r}
#statistical analysis for age and education
adult %>% group_by(income) %>% dplyr::summarize(N = n(),
                                                mean_of_age = mean(age),
                                                strd_dev_of_age = sd(age),
                                                mean_of_education_num = mean(education_num),
                                                std_dev_of_education_num = sd(education_num),
                                                correlation_between_x_and_y = cor(age,education_num))

#statistical analysis for log age and education_nume
adult %>% group_by(income) %>% dplyr::summarize(N = n(),
                                                mean_of_age = mean(age),
                                                strd_dev_of_age = sd(age),
                                                mean_of_education_num = mean(education_num),
                                                std_dev_of_education_num = sd(education_num),
                                                correlation_between_x_and_y = cor(log(age),log(education_num)))



#histograms of income by marital status
ggplot(adult, aes(x=as.numeric(age), group = marital_status, fill = marital_status)) + 
  geom_histogram(binwidth = 5, color = "black") + 
  xlab("Marital Status Population of Census 1994") + 
  ggtitle("Histogram of income") + 
  theme(plot.title = element_text(hjust = 0.5))

#facet wrap histogram of age by marital status
ggplot(adult, aes(x=as.numeric(age), group = marital_status, fill = marital_status)) + 
  geom_histogram(binwidth = 5, color = "black") + 
  xlab("Marital Status Population of Census 1994") + 
  ggtitle("Histogram of income") + 
  theme(plot.title = element_text(hjust = 0.5))+
  facet_wrap(~marital_status)

#histograms of income by age
ggplot(adult, aes(x=as.numeric(age), group = sex, fill = sex)) + geom_histogram(binwidth = 5, color = "black") + xlab("Sex Population of Census 1994") + ggtitle("Histogram of sex") + theme(plot.title = element_text(hjust = 0.5))


#histograms of income by age
ggplot(adult, aes(x=as.numeric(age), group = income, fill = income)) + geom_histogram(binwidth = 5, color = "black") + xlab("Age Population of Census 1994") + ggtitle("Histogram of income") + theme(plot.title = element_text(hjust = 0.5))

#statistics based on occupation
adult_occup <- adult %>% group_by(occupation, income) %>% dplyr::summarise(N = n()) %>%
  mutate(Freq = round(N/sum(N),3))

#print out the adult_occup we just created
adult_occup


#breakdown of salaries based on jobs barplot
adult_occup %>% ggplot(aes(occupation, y = N)) +
  geom_bar(aes(fill = income), stat = "identity")

#statistics based on sex
adult_sex <- adult %>% group_by(sex, income) %>% dplyr::summarise(N = n()) %>%
  mutate(Freq = round(N/sum(N),3))

#print out the adult_sex we just created
adult_sex

#breakdown of salaries based on sex barplot
adult_sex %>% ggplot(aes(sex, y = N)) +
  geom_bar(aes(fill = income), stat = "identity") + 
  ggtitle("Barplot by sex and income") + 
  theme(plot.title = element_text(hjust = 0.5))


#breakdown of salaries based on continent barplot
adult_cont <- adult %>% group_by(native_country, income) %>% dplyr::summarise(N = n()) %>%
  mutate(Freq = round(N/sum(N),3))

#print out the adult_cont we just created
adult_cont

#breakdown of salaries based on native continent barplot
adult_cont %>% ggplot(aes(native_country, y = N)) +
  geom_bar(aes(fill = income), stat = "identity") + xlab("Native Continent") +
  ggtitle("Barplot by native continent and income") + 
  theme(plot.title = element_text(hjust = 0.5))
  
  
#breakdown of salaries based on workclass barplot
adult_work <- adult %>% group_by(workclass, income) %>% dplyr::summarise(N = n()) %>%
  mutate(Freq = round(N/sum(N),3))

#print out the adult_work we just created
adult_work

#breakdown of salaries based on native workclass barplot
adult_work %>% ggplot(aes(workclass, y = N)) +
  geom_bar(aes(fill = income), stat = "identity") + xlab("Workclass") +
  ggtitle("Barplot by Workclass and income") + 
  theme(plot.title = element_text(hjust = 0.5))  


adult_num <- dplyr::select(adult,age,education_num,capital_gain,capital_loss,hours_per_week,income)

view(adult_num)

pairs(adult_num[1:5])

#transformation data
adult_num$age <- log(adult_num$age)
adult_num$education <- log(adult_num$education)
adult_num$capital_gain <- log(adult_num$capital_gain)
adult_num$capital_loss <- log(adult_num$capital_loss)
adult_num$hours_per_week <- log(adult_num$hours_per_week)

#transforming the pairs
pairs(adult_num[1:5])

#correlation testing between variables
cor(adult_num$age,adult_num$education)
cor(adult_num$age,adult_num$capital_gain)
cor(adult_num$capital_loss,adult_num$age)
cor(adult_num$age,adult_num$hours_per_week)
```




```{r}
#reviewing factors
levels(adult$marital_status)

#Converting to factors
adult$workclass <- as.numeric(as.factor(adult$workclass))
adult$marital_status <- as.numeric(as.factor(adult$marital_status))
adult$race <- as.numeric(as.factor(adult$race))
adult$sex <- as.numeric(as.factor(adult$sex))
adult$native_country <- as.numeric(as.factor(adult$native_country))
adult$income <- as.factor(adult$income)

#reviewing factors
levels(adult$workclass)
levels(adult$marital_status)
levels(adult$race)
levels(adult$sex)
levels(adult$native_country)
levels(adult$income)
                   
               
#renaming the native country column to continent
adult <- rename(adult, "continent" = "native_country")

#factorial analysis
table(adult$race, adult$income)
table(adult$continent, adult$income)
table(adult$marital_status,adult$income)
table(adult$sex,adult$income)
table(adult$workclass,adult$income)

```
finding out the outliers and dropping them
```{r}

#removing outliers in the dataset
adult <- adult %>% mutate(is_outlier = age <= 65) %>% filter(is_outlier == TRUE) %>% dplyr::select(-is_outlier)

adult <- adult %>% mutate(is_outlier = education_num > 4) %>% filter(is_outlier == TRUE) %>% dplyr::select(-is_outlier)

adult <- adult %>% mutate(is_outlier = hours_per_week >= 38 & hours_per_week <= 42) %>% filter(is_outlier == TRUE) %>% dplyr::select(-is_outlier)



ggplot(adult, aes(age)) + 
  geom_boxplot()+coord_flip() +
  ggtitle("Age Boxplot ") +
  theme(plot.title = element_text(hjust = 0.5))

ggplot(adult, aes(education_num)) + 
  geom_boxplot()+coord_flip() +
  ggtitle("Education_num Boxplot ") +
  theme(plot.title = element_text(hjust = 0.5))

ggplot(adult, aes(hours_per_week)) + 
  geom_boxplot()+coord_flip() +
  ggtitle("Hours_per_week Boxplot ") +
  theme(plot.title = element_text(hjust = 0.5))

```






```{r}
#splitting the data into train and test

ratio<-sample(1:nrow(adult), size = 0.25 * nrow(adult))
train_data <- adult[-ratio,]
test_data <- adult[ratio,]
dim(train_data)
dim(test_data)

head(train_data)


```
objective 1 build a simple logistic model

```{r}

library(ROCR)

#simple logistic regression
logreg1 <- glm(income ~., data = train_data, family = binomial('logit'))
summary(logreg1)

confint(logreg1)

summary(logreg1)

full.model <- logreg1
null.model <- glm(income ~ 1, data = train_data, family = binomial('logit'))

#backward selection model
step(full.model, trace = F, scope = list(lower=formula(null.model), upper=formula(full.model)),direction = 'backward')

#forward selection model
step(null.model, trace = F, scope = list(lower=formula(null.model), upper=formula(full.model)),direction = 'forward')

#stepwise selection model
#forward selection model
step(null.model, trace = F, scope = list(lower=formula(null.model), upper=formula(full.model)),direction = 'both')

#check accuracy
predict1 <- predict(logreg1, train_data, type = "response")
ROC_pred <- prediction(predict1, train_data$income)
ROC_perf <- performance(ROC_pred, "tpr", "fpr")
plot(ROC_perf, colorize=TRUE, print.cutoffs.at = seq(0,1,0.1), text.adj = c(-0.2, 1.7))
abline(a=0, b=1)
auc_train <- round(as.numeric(performance(ROC_pred, "auc")@y.values),2)
legend(.8, .2, auc_train, title = "AUC", cex=1)

#make predictions on test data
Predict_test <- predict(logreg1, type = "response", newdata = test_data)
test_tab <- table(test_data$income, Predict_test > 0.5)
test_tab

#residual plots
logreg_summary <- broom::augment(logreg1) %>% mutate(index = 1:n())

logreg_summary %>% top_n(3, .cooksd)

ggplot(logreg_summary, aes(index, .std.resid)) + 
  geom_point(aes(color = income), alpha = 0.5) + 
  theme_bw() 

#cooks D checking assumptions
plot(logreg1, pch = 18, col = "red", which = c(6))

car::vif(logreg1)

#confidence interval
CI_lower <- coefficients(logreg1)[2] - 1.96*summary(logreg1)$coefficients[2,2]
CI_upper <- coefficients(logreg1)[2] + 1.96*summary(logreg1)$coefficients[2,2]

CI_lower
CI_upper

```




objective 2 build a complex logistic model
```{r}
#complex logistic regression
logreg2 <- glm(income ~ education_num + capital_gain + 
    + log(age) + log(hours_per_week) + capital_loss +  
    education_num: capital_gain + age:hours_per_week, data = train_data, family = binomial('logit'))
summary(logreg2)

full.model2 <- logreg2
null.model2 <- glm(income ~ 1, data = train_data, family = binomial('logit'))

#backward selection model
step(full.model2, trace = F, scope = list(lower=formula(null.model2), upper=formula(full.model2)),direction = 'backward')

#forward selection model
step(null.model2, trace = F, scope = list(lower=formula(null.model2), upper=formula(full.model2)),direction = 'forward')

#stepwise selection model
step(null.model2, trace = F, scope = list(lower=formula(null.model2), upper=formula(full.model2)),direction = 'both')

#check accuracy
predict2 <- predict(logreg2, train_data, type = "response")
ROC_pred2 <- prediction(predict2, train_data$income)
ROC_perf2 <- performance(ROC_pred2, "tpr", "fpr")
plot(ROC_perf2, colorize=TRUE, print.cutoffs.at = seq(0,1,0.1), text.adj = c(-0.2, 1.7))
abline(a=0, b=1)
auc_train2 <- round(as.numeric(performance(ROC_pred2, "auc")@y.values),2)
legend(.8, .2, auc_train2, title = "AUC", cex=1)

#make predictions on test data
Predict_test2 <- predict(logreg2, type = "response", newdata = test_data)
test_tab2 <- table(test_data$income, Predict_test2 > 0.5)
test_tab2

accuracy_test2 <- round(sum(diag(test_tab2))/sum(test_tab2),2)
sprintf("Accuracy on test set is %s", accuracy_test2)

prob <- predict(logreg2, test_data, type = 'response')

#residual plots
logreg_summary2 <- broom::augment(logreg2) %>% mutate(index = 1:n())

logreg_summary %>% top_n(3, .cooksd)

ggplot(logreg_summary2, aes(index, .std.resid)) + 
  geom_point(aes(color = income), alpha = 0.5) + 
  theme_bw() 

#cooks D checking assumptions
plot(logreg2, pch = 18, col = "red", which = c(6))

car::vif(logreg2)

#confidence interval
CI_lower2 <- coefficients(logreg2)[2] - 1.96*summary(logreg1)$coefficients[2,2]
CI_upper2 <- coefficients(logreg2)[2] + 1.96*summary(logreg1)$coefficients[2,2]

CI_lower2
CI_upper2



```



object 2 LDA model
```{r}
	# LDA 
#Linear Discriminate Analysis, overall, resulted in lower accuracy over logistic regression. The most probable reason #for this is because linear discriminate analysis assumes normally distributed predictors while logistic regression #does not

#LDA with all predictors, Accuracy 83.31%
ldafit <- lda(income~., data=train_data)
ldapred <- predict(ldafit, newdata=test_data, type = 'response')
ldapred2 <- predict(ldafit, newdata=test_data, type = 'class')
ldatable <- table(ldapred$class, test_data$income)
lda.acc <- mean(ldapred$class==test_data$income)
lda.acc

# LDA excluding race and sex  acuuracy little lower: 83.19%
ldafit <- lda(income~.-race-sex, data=train_data)
ldapred <- predict(ldafit, newdata=test_data)
ldatable <- table(ldapred$class, test_data$income)
lda.acc2 <- mean(ldapred$class==test_data$income)
lda.acc2

```



Machine learning models for Objective 2 trees and random forest
```{r}
library(rpart)
#decision trees model
trees <- rpart(income ~., data=train_data, method = 'class', cp = 1e-3)
trees_pred_prob <- predict(trees, newdata = test_data, type = 'prob')
trees_pred <- predict(trees, newdata = test_data, type = 'class')

#confusion matrix accuracy was 82.94%
trees_cm <- table(trees_pred, test_data$income)
trees_cm

confusionMatrix(trees_pred, test_data$income)

library(randomForest)
#random forest model
rf_model <- randomForest(income ~., data = train_data, ntree = 950)
rf_model_predprob <- predict(rf_model, newdata = test_data, type = 'prob')
rf_model_pred <- predict(rf_model, newdata = test_data, type = 'class')

#confusion matrix, random forest model has accuracy of 86.19%
rf_cm <- table(rf_model_pred, test_data$income)
rf_cm

confusionMatrix(rf_model_pred, test_data$income)

```

ROC curve calculation
```{r}
# create simple logistic regression
pr <- prediction(prob, test_data$income)
prf <- performance(pr, measure = "tpr", x.measure = "fpr")
dd <- data.frame(FP = prf@x.values[[1]], TP = prf@y.values[[1]])

#complex logistic regression
pr1 <- prediction(prob, test_data$income)
prf1 <- performance(pr1, measure = "tpr", x.measure = "fpr")
dd1 <- data.frame(FP = prf@x.values[[1]], TP = prf1@y.values[[1]])

#Decision Trees model
pr2 <- prediction(trees_pred_prob[,2], test_data$income)
prf2 <- performance(pr2, measure = "tpr", x.measure = "fpr")
dd2 <- data.frame(FP = prf2@x.values[[1]], TP = prf2@y.values[[1]])


# Random forest model
pr3 <- prediction(rf_model_predprob[,2], test_data$income)
prf3 <- performance(pr3, measure = "tpr", x.measure = "fpr")
dd3 <- data.frame(FP = prf3@x.values[[1]], TP = prf3@y.values[[1]])


# plot ROC curve for logistic regression
g <- ggplot() + 
  geom_line(data = dd, aes(x = FP, y = TP, color = 'Simple Logistic Regression')) +
  geom_line(data = dd1, aes(x = FP, y = TP, color = 'Complex Logistic Regression')) +
  geom_line(data = dd2, aes(x = FP, y = TP, color = 'Decision Trees')) +
  geom_line(data = dd3, aes(x = FP, y = TP, color = 'Random Forest')) +
  geom_segment(aes(x = 0, xend = 1, y = 0, yend = 1)) +
  ggtitle('ROC Curve') + 
  labs(x = 'False Positive Rate', y = 'True Positive Rate') 


g +  scale_colour_manual(name = 'Classifier', values = c('Simple Logistic Regression'='#E69F00', 
                                               'Complex Logistic Regression'='#56B4E9', 'Decision Trees'='#009E73', 
                                               'Random Forest'='#D55E00')) + theme(plot.title = element_text(hjust = 0.5))

```







